---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# nestedscales

<!-- badges: start -->
<!-- badges: end -->

nestedscales represents a discrete data dimension at multiple resolutions
and recasts tabular data between them. A **nested scale** stores one row per
finest-grained unit, or *atom*, and one column for each coarser grouping, or
*frame*. Each frame assigns atoms to units; frames may form a hierarchy or
cross-cut one another, including product-like overlays. All mappings are
derived from membership at the atom level.

The package defines:

* **Structure**: the levels of a scale, the units at each level, and the
  relations between levels (nested, crossed, uniform).
* **Transition rules** between levels, chosen per value column and aware of
  direction: `sum`, `weighted_mean`, `mean`, and `copy`; `sd` for
  aggregation; and `share` or `logshare` for within-parent shares.
* **Operations**: converting data between levels, attaching level labels to
  a dataset, subsetting, combining several scales into one multi-dimensional
  index, clustering units into new levels, and storing data together with
  its scales.

The package is dimension-agnostic. Time and space are provided by two
packages built on it, designed for optimization and simulation models, where
labeled time slices and regions serve as the index sets of variables and
constraints:

* [timescales](https://github.com/optimal2050/timescales): `Calendar`,
  labeled discrete time slices (hours, days, seasons, representative days).
* [geoscales](https://github.com/optimal2050/geoscales): `Geoscale`,
  labeled geographical regions with geometry.

Both classes are subclasses of `NestedScale`. Any other dimension
(industries, income brackets, technology vintages) is declared directly from
a table. `vignette("nestedscales")` introduces the functions; the
[package website](https://optimal2050.github.io/nestedscales/) includes a
roadmap of what is available and what is planned.

## Installation

Until the first CRAN release, install the development version from GitHub:

```r
pak::pkg_install("optimal2050/nestedscales")
```

The package is pure R; its non-base runtime dependencies are dplyr, rlang
and S7.

## A scale in one call

Declare the levels, coarsest first:

```{r example}
library(nestedscales)

industries <- data.frame(
  section  = c("C", "C", "C", "D", "D"),
  division = c("C10", "C10", "C11", "D35", "D35"),
  class    = c("C101", "C102", "C110", "D351", "D352"),
  gva      = c(120, 80, 45, 300, 150)
)

ind <- scale_from_leaftable(
  industries,
  frames = c("section", "division", "class"),
  weights = "gva",
  name   = "nace"
)

ind
```

Every structural question is answered from the atoms, never from an assumed
tree:

```{r queries}
scale_frames(ind)
scale_units(ind, "division")
scale_nests(ind, "section", "division")
```

Hierarchies that cross-cut are normal and supported -- one detailed code may
belong to several aggregates. `scale_nests(x, parent, child)` tests whether
every child code belongs to a single parent code; conversion never depends on
that relationship.

```{r summary}
summary(ind)
```

## Moving data between levels

Data keyed at one level is recast to another with a rule per value column.
With complete coverage, `"sum"` conserves totals; intensive quantities use
`"weighted_mean"` with a declared weight:

```{r recast}
x <- data.frame(
  class = c("C101", "C102", "C110", "D351", "D352"),
  jobs  = c(12, 8, 5, 30, 20),
  wage  = c(31, 29, 27, 45, 40)
)

recast_scale(x, ind, from = "class", to = "section",
             rule = c(jobs = "sum", wage = "weighted_mean"),
             weight = "gva")
```

## Design

* **Atoms are the ground truth.** Frames are partitions of the atoms, not
  necessarily a strict tree; conversion always routes through the atom layer.
* **Weights are named and plural.** A scale can carry `gva`, `employment` and
  `count` side by side, with one declared default.
* **Partial coverage is first class.** An atom with no code at a frame is
  `NA`, not an error, and a subset records what fraction of its parent it
  keeps.
* **Backend-agnostic.** The verbs use the dplyr interface, so the same
  pipeline runs on a `data.frame`, a `data.table`, or an arrow dataset;
  lazy inputs stay lazy.

## Related packages

* [scales](https://CRAN.R-project.org/package=scales) maps data to
  graphical aesthetics; its discrete scales are palettes and label
  formatters, not resolutions of a model dimension.
* [data.tree](https://CRAN.R-project.org/package=data.tree) represents
  strict trees as node objects. nestedscales keeps the hierarchy as a flat
  atom table, allows frames that cross-cut, and converts data between any
  two frames with weights.
* [hts](https://CRAN.R-project.org/package=hts) reconciles forecasts across a
  fixed aggregation tree; `reconcile_scale()` here compares data aggregated
  with `sum` or `weighted_mean` against independent totals and can balance
  discrepancies.

## License

Apache License 2.0.
