---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# nestedscales

<!-- badges: start -->
<!-- badges: end -->

Nested scales for optimization and simulation models, without naming the
dimension. A **scale** is a flat table of *atoms* plus the *frames* that
group them. Every frame nests in the atoms; frames need not nest in each
other, so overlapping layers are ordinary. That shape fits time, space,
industries, income brackets, temperature regimes or technology vintages
equally well.

`nestedscales` is the engine under [timescales](https://github.com/optimal2050/timescales)
(calendars) and [geoscales](https://github.com/optimal2050/geoscales)
(regions), whose classes are subclasses of `NestedScale`. Use it directly for any
other dimension.

## Installation

From CRAN:

```r
install.packages("nestedscales")
```

The development version from GitHub:

```r
pak::pkg_install("optimal2050/nestedscales")
```

The package is pure R; its hard dependencies are dplyr, rlang and S7.

## A scale in one call

Declare the hierarchy, coarsest frame first:

```{r example}
library(nestedscales)

industries <- data.frame(
  section  = c("C", "C", "C", "D", "D"),
  division = c("C10", "C10", "C11", "D35", "D35"),
  class    = c("C101", "C102", "C110", "D351", "D352"),
  gva      = c(120, 80, 45, 300, 150)
)

ind <- scale_from_leaftable(
  industries,
  frames = c("section", "division", "class"),
  name   = "nace"
)

ind
```

Every structural question is answered from the atoms, never from an assumed
tree:

```{r queries}
scale_frames(ind)
scale_units(ind, "division")
scale_nests(ind, "section", "division")
```

Hierarchies that cross-cut are normal and supported — one detailed code may
belong to several aggregates. `scale_nests()` tells you whether a given pair
happens to nest; conversion never depends on it.

```{r summary}
summary(ind)
```

## Design

* **Atoms are the ground truth.** Frames are partitions of the atoms, not
  necessarily a strict tree; conversion always routes through the atom layer.
* **Weights are named and plural.** A scale can carry `gva`, `employment` and
  `count` side by side, with one declared default.
* **Partial coverage is first class.** An atom with no code at a frame is
  `NA`, not an error, and a subset records what fraction of its parent it
  keeps.
* **Backend-agnostic.** The verbs are written on dplyr only, so the same
  pipeline runs on a `data.frame`, a `data.table`, or an arrow dataset;
  lazy inputs stay lazy.

## Related packages

* [scales](https://CRAN.R-project.org/package=scales) maps data to
  graphical aesthetics; its discrete scales are palettes and label
  formatters, not resolutions of a model dimension.
* [data.tree](https://CRAN.R-project.org/package=data.tree) represents
  strict trees as node objects. nestedscales keeps the hierarchy as a flat
  atom table, allows frames that cross-cut, and converts data between any
  two frames with weights.
* [hts](https://CRAN.R-project.org/package=hts) reconciles forecasts across a
  fixed aggregation tree; `reconcile_scale()` here balances observed data
  against a total, for any scale and rule.

## License

Apache License 2.0.
