---
title: "Writing a dimension package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing a dimension package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# a twin built against an older core is not a DiscreteScale here; skip it as absent
has_ts <- requireNamespace("timescales", quietly = TRUE) &&
  discretescales::scale_is(timescales::calendar("m12_h24"))
has_gs <- requireNamespace("geoscales", quietly = TRUE) &&
  discretescales::scale_is(geoscales::geoscale_example())
```

## What discretescales is for

Models rarely use data at the resolution it arrives in. Demand comes by hour
and the model runs on twelve representative days; statistics come by
district and the model has five regions; accounts come by detailed industry
and the model has six sectors. Each of these is a *scale*: a set of fine
units, the **atoms**, grouped into coarser levels, the **frames**. Hours sit
inside days inside months; districts inside provinces inside countries.

discretescales stores such a hierarchy as a `DiscreteScale` and converts data along it:
summing extensive quantities, averaging intensive ones with the right
weights, splitting coarse values down to fine units, and attaching group
labels to a dataset. The same functions work whatever the dimension is,
because a scale is just a table of atoms with one column per frame.

```{r setup}
library(discretescales)
```

## Dimension packages

Some dimensions need more than a table. Time has calendars, leap years,
fiscal years, and a mapping from clock time to the model's time slices.
Space has maps, geometry and coordinate systems. These live in **dimension
packages** built on discretescales:

* **timescales** provides `Calendar`, a scale of time: a catalog of standard
  calendars, conversion from datetimes, calendar plots.
* **geoscales** provides `Geoscale`, a scale of space: region hierarchies
  from data providers, geometry, maps.

Both classes are subclasses of `DiscreteScale`, so each package adds what is special
about its dimension and gets the conversion machinery from discretescales. A
`Calendar` is a `DiscreteScale`:

```{r, eval = has_ts, purl = FALSE}
cal <- timescales::calendar("m12_h24") # 12 months x 24 hours
scale_is(cal)
scale_frames(cal)
head(scale_leaftable(cal), 3)
```

and so is a `Geoscale`:

```{r, eval = has_gs, purl = FALSE}
gs <- geoscales::geoscale_example()
scale_is(gs)
scale_frames(gs)
head(scale_leaftable(gs), 3)
```

so the functions of this package work on either. Hourly load summed to
months uses the same `recast_scale()` as any other scale:

```{r, eval = has_ts, purl = FALSE}
load <- data.frame(timeslice = scale_units(cal), mw = 1)
head(recast_scale(load, cal, from = "timeslice", to = "MONTH", rule = "sum"), 3)
```

and the regions of country `N` come from the same navigation function as
anything else:

```{r, eval = has_gs, purl = FALSE}
scale_children(gs, "country", "N")
```

The dimension packages wrap these in their own vocabulary --
`timescales::recast_calendar()`, `geoscales::geoscale_children()` -- but the
work underneath is shared.

## Do you need a package?

Often not. Any hierarchy that fits in a table can be used directly:

```{r}
sectors <- data.frame(
  section  = c("C", "C", "C", "D", "D"),
  division = c("C10", "C10", "C24", "D35", "D35"),
  activity = c("C10.1", "C10.5", "C24.1", "D35.1", "D35.2"),
  gva      = c(30, 20, 50, 80, 20)
)
nace <- scale_from_leaftable(sectors,
  frames = c("section", "division", "activity"),
  key = "activity", name = "nace"
)
nace
```

Write a dimension package when the dimension has something of its own to
offer: standard classifications to ship, its own way of building scales,
extra data per atom, plots, or simply a vocabulary its users expect. The rest
of this article builds a small one for industry classifications, and shows
at each step how timescales and geoscales do the same thing.

## Step 1: a subclass

A dimension class is an S7 class whose parent is `scale_class()`. (The
`DiscreteScale` class itself is reached this way rather than exported; the
package exports functions only.)

Two parts are typical. A **constructor** fixes what is fixed for the
dimension, such as the name of the key column and the weights, and passes
the rest to `scale_from_leaftable()`. An **alias property** lets users call
the frames by the dimension's own word; `scale_alias_property()` builds it.

The industry class calls its frames `levels`, and also keeps a title for
each activity:

```{r}
titles <- data.frame(
  activity = sectors$activity,
  title = c(
    "Meat processing", "Dairy products", "Basic iron and steel",
    "Electric power generation", "Gas distribution"
  )
)

Industry <- S7::new_class(
  "Industry",
  parent = scale_class(),
  properties = list(
    levels = scale_alias_property("frames"),
    titles = S7::class_data.frame
  ),
  constructor = function(leaftable, levels, titles) {
    s <- scale_from_leaftable(leaftable,
      frames = levels, key = "activity",
      weights = "gva", name = "nace"
    )
    S7::new_object(s, titles = titles)
  },
  validator = function(self) {
    if (!all(self@leaftable$activity %in% self@titles$activity)) {
      "every activity needs a title"
    }
  }
)

ind <- Industry(sectors,
  levels = c("section", "division", "activity"),
  titles = titles
)
ind@levels
```

The two dimension packages follow the same pattern. A `Calendar` calls its
frames `timeframes` and always keys its atoms by `timeslice`; a `Geoscale`
calls them `geoframes`, keys them by `region`, and carries geometry:

```{r, eval = FALSE}
# timescales
Calendar <- S7::new_class(
  "Calendar",
  parent = discretescales::scale_class(),
  properties = list(timeframes = discretescales::scale_alias_property("frames")),
  ...
)

# geoscales
Geoscale <- S7::new_class(
  "Geoscale",
  parent = discretescales::scale_class(),
  properties = list(
    geoframes = discretescales::scale_alias_property("frames"),
    geometry  = S7::new_property(S7::class_any, default = NULL)
  ),
  ...
)
```

## Step 2: the shared functions already work

Nothing more is needed for the industry scale to convert data:

```{r}
jobs <- data.frame(activity = sectors$activity, jobs = c(5, 3, 8, 2, 1))
recast_scale(jobs, ind,
  from = "activity", to = "section",
  values = "jobs", rule = "sum"
)
```

The same holds for `join_scale()`, `filter_scale()`, the navigation
functions, `scale_product()`, clustering and the Arrow store. The generic
`recast()` dispatches on the new class too:

```{r}
recast(jobs, ind, to = "division", values = "jobs", rule = "sum")
```

## Step 3: the dimension's vocabulary

Error messages from the shared functions are written once and filled in
with words from `scale_vocab()`. Without a method, an industry scale is
described in neutral words:

```{r, error = TRUE}
recast_scale(jobs, ind,
  from = "activity", to = "group",
  values = "jobs", rule = "sum"
)
```

A method supplies the dimension's own:

```{r, error = TRUE}
S7::method(scale_vocab, Industry) <- function(x, ...) {
  list(
    object = "Industry", frame = "level", frames = "levels",
    unit = "industry", units = "industries", atoms = "activities"
  )
}

recast_scale(jobs, ind,
  from = "activity", to = "group",
  values = "jobs", rule = "sum"
)
```

This is why a calendar's errors talk about timeframes and a region
hierarchy's about geoframes:

```{r, eval = has_ts, error = TRUE, purl = FALSE}
scale_vocab(cal)$frame
recast_scale(load, cal, from = "timeslice", to = "WEEK", rule = "sum")
```

The vocabulary may also name the dimension package's own registration
functions (`register_rule`, `register_map`), so that a message suggesting
one points to, say, `register_geoscale_rule()` rather than the generic
`register_scale_rule()`.

## Step 4: data kept per atom

When a function keeps only some atoms -- `filter_scale()`, `prune_scale()`,
`x[frame, unit]` -- the result keeps its class and any extra properties.
If those properties hold one entry per atom, a `scale_payload_slice()`
method cuts them down to match:

```{r}
S7::method(scale_payload_slice, Industry) <- function(x, i, ...) {
  x@titles <- x@titles[x@titles$activity %in% x@leaftable$activity, ,
    drop = FALSE
  ]
  x
}

manufacturing <- filter_scale(ind, "section", "C")
class(manufacturing)
manufacturing@titles
scale_coverage(manufacturing)
```

geoscales uses the same hook for its geometry, which has one shape per atom
in leaftable order, so a filtered Geoscale keeps exactly the shapes of the
regions it kept. The object is validated once the hook has run, so the data
may be positional, like geometry, or keyed by atom code, like the titles
here.

## Atoms that are combinations of frames

In the industry scale every activity belongs to one division and every
division to one section: the frames nest, and the finest frame lists the
atoms. Time does not work that way. In `m12_h24` the frames are months and
hours of the day; an hour of the day is not inside a month, and the atoms --
288 time slices -- are combinations of the two. discretescales handles this by
taking the atoms from the key column when no frame lists them.
`scale_atom_level()` reports which case applies:

```{r}
scale_atom_level(ind)
```

```{r, eval = has_ts, purl = FALSE}
scale_atom_level(cal)
length(scale_units(cal))
```

A dimension whose atoms are combinations of other categories works the same
way, with no extra code.

## Turning it into a package

Inside a script, methods can be registered directly, as above. A package
registers them on the discretescales generics with S7's external generics, so
that discretescales finds them without depending on the package:

```{r, eval = FALSE}
# R/industry-class.R
.scale_vocab_generic <- S7::new_external_generic(
  "discretescales",
  "scale_vocab", "x"
)
S7::method(.scale_vocab_generic, Industry) <- function(x, ...) {
  list(
    object = "Industry", frame = "level", frames = "levels",
    unit = "industry", units = "industries", atoms = "activities"
  )
}

# R/zzz.R
.onLoad <- function(libname, pkgname) {
  S7::methods_register()
}
```

The DESCRIPTION names discretescales and S7:

```
Imports:
    discretescales,
    S7
```

Both dimension packages also give `recast()` a method of their own, because
their conversions take extra arguments -- timescales needs a `year` to place
a calendar on real dates -- and re-export the generic so that
`library(timescales)` or `library(geoscales)` alone provides it:

```{r, eval = FALSE}
.recast_generic <- S7::new_external_generic(
  "discretescales", "recast",
  c("x", "from")
)
S7::method(.recast_generic, list(S7::class_any, Industry)) <-
  function(x, from, to, ...) {
    recast_scale(x, from, to = to, ...)
  }

#' @importFrom discretescales recast
#' @export
discretescales::recast
```

When a dimension's conversions cannot be read off one scale's table --
timescales maps one calendar onto another through a grid of real dates --
the package builds the correspondence itself and hands it to
`recast_crosswalk()`, which applies the same rules as `recast_scale()`.

## Things to keep in mind

* **Test class membership with `scale_is()`, not by name.** Inside an
  installed package an S7 class is called `"yourpackage::Industry"`, but
  under `pkgload::load_all()` it is plain `"Industry"`.
* **Saved objects carry their class.** An object saved to `data/` or an
  `.rds` file records its class and parent. If either changes, the saved
  object stops being a `DiscreteScale` and the shared functions reject it with
  "`x` must be a DiscreteScale object". Rebuild saved objects after changing a class.
* **Keep dimension-specific code in the dimension package.** discretescales
  depends on no dimension package, and nothing in it needs to know that
  calendars or regions exist.
