---
title: "Writing a dimension package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing a dimension package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

multiscales knows nothing about any particular dimension. A dimension --
time, space, an industry classification, an income schedule -- lives in its
own package, which defines a subclass of the multiscales `Scale` and adds what
is specific to it: catalogs, data, plots, articles. Every multiscales verb
then works on the new class unchanged, because the new class *is* a Scale.

[timescales](https://github.com/optimal2050/timescales) (`Calendar`) and
[geoscales](https://github.com/optimal2050/geoscales) (`Geoscale`) are built
this way. This article walks through a small one: an industry classification.

```{r setup}
library(multiscales)
```

## The data

A dimension starts from a leaftable: one row per atom, one column per frame,
coarsest first, plus any numeric weight columns.

```{r}
nace <- data.frame(
  section  = c("C", "C", "C", "D", "D"),
  division = c("C10", "C10", "C24", "D35", "D35"),
  activity = c("C10.1", "C10.5", "C24.1", "D35.1", "D35.2"),
  gva      = c(30, 20, 50, 80, 20)
)
```

## The subclass

The class is an S7 class whose parent is `scale_class()`. The `Scale` class
itself is not exported -- its name would mask `ggplot2::Scale` -- so this is
the way to reach it.

Two things are typical:

* an **alias property** that names the inherited `frames` in the dimension's
  own words (`timeframes` for a Calendar, `geoframes` for a Geoscale, `levels`
  here), built with `scale_alias_property()`;
* a **constructor** that fixes what is fixed for the dimension -- the key
  column, the weight columns -- and delegates the rest to
  `scale_from_leaftable()`.

A subclass may also carry data per atom. This one keeps a title for each
activity:

```{r}
titles <- data.frame(
  activity = nace$activity,
  title = c("Meat processing", "Dairy products", "Basic iron and steel",
            "Electric power generation", "Gas distribution")
)

Industry <- S7::new_class(
  "Industry",
  parent = scale_class(),
  properties = list(
    levels = scale_alias_property("frames"),
    titles = S7::class_data.frame
  ),
  constructor = function(leaftable, levels, titles) {
    s <- scale_from_leaftable(leaftable, frames = levels, key = "activity",
                              weights = "gva", name = "nace")
    S7::new_object(s, titles = titles)
  },
  validator = function(self) {
    if (!all(self@leaftable$activity %in% self@titles$activity)) {
      "every activity needs a title"
    }
  }
)

ind <- Industry(nace, levels = c("section", "division", "activity"),
                titles = titles)
ind@levels
```

## Everything already works

No method has been written yet, and the whole engine is available:

```{r}
scale_is(ind)
scale_units(ind, "division")

jobs <- data.frame(activity = nace$activity, jobs = c(5, 3, 8, 2, 1))
recast_scale(jobs, ind, from = "activity", to = "section",
             values = "jobs", rule = "sum")
```

The bare `recast()` generic dispatches on the Scale method too:

```{r}
recast(jobs, ind, to = "division", values = "jobs", rule = "sum")
```

So do `join_scale()`, `filter_scale()`, the navigation verbs,
`scale_product()`, clustering and the Arrow store.

## Speaking the dimension's language

The engine's messages are written once and filled in from `scale_vocab()`.
Without a method an Industry is described in neutral words:

```{r, error = TRUE}
recast_scale(jobs, ind, from = "activity", to = "group",
             values = "jobs", rule = "sum")
```

A method supplies the dimension's own:

```{r, error = TRUE}
S7::method(scale_vocab, Industry) <- function(x, ...) {
  list(object = "Industry", frame = "level", frames = "levels",
       unit = "industry", units = "industries", atoms = "activities")
}

recast_scale(jobs, ind, from = "activity", to = "group",
             values = "jobs", rule = "sum")
```

## Carrying data per atom

When a verb keeps only some atoms -- `filter_scale()`, `prune_scale()`, `[` --
the engine rebuilds the object from a copy, so the class and every extra
property survive. It then calls `scale_payload_slice()` so the subclass can cut
its own per-atom data down to match:

```{r}
S7::method(scale_payload_slice, Industry) <- function(x, i, ...) {
  x@titles <- x@titles[x@titles$activity %in% x@leaftable$activity, ,
                       drop = FALSE]
  x
}

mfg <- filter_scale(ind, "section", "C")
class(mfg)
mfg@titles
scale_coverage(mfg)
```

The engine validates the object once, after the hook has run, so the payload
may be positional -- one element per leaftable row, as geoscales' geometry is
-- or keyed by atom code, as the titles are here.

## Frames that cross-cut

The industry frames nest: every activity sits in one division, every division
in one section. Not every dimension is like that. A calendar's `YDAY` and
`HOUR` cut across each other, and its atoms -- the 8760 timeslices -- are
combinations of the two rather than codes of either. multiscales handles this
when the key column holds the atoms: `scale_atom_level()` then returns the key
instead of the finest frame, and the verbs route through it.

```{r}
scale_atom_level(ind)
```

## In a package

A script can register methods directly, as above. A package registers them on
the multiscales generics through S7's external-generic mechanism, so that they
are found without multiscales importing the dimension package:

```{r, eval = FALSE}
# R/industry-class.R
.scale_vocab_generic <- S7::new_external_generic("multiscales",
                                                 "scale_vocab", "x")
S7::method(.scale_vocab_generic, Industry) <- function(x, ...) {
  list(object = "Industry", frame = "level", frames = "levels",
       unit = "industry", units = "industries", atoms = "activities")
}

# A dimension-specific recast method, if the dimension needs extra arguments
# (timescales' takes `year`, for instance):
.recast_generic <- S7::new_external_generic("multiscales", "recast",
                                            c("x", "from"))
S7::method(.recast_generic, list(S7::class_any, Industry)) <-
  function(x, from, to, ...) {
    recast_industry(x, from, to, ...)
  }

# R/zzz.R
.onLoad <- function(libname, pkgname) {
  S7::methods_register()
}
```

and declares the dependency:

```
Imports:
    multiscales,
    S7
```

Re-exporting the generic, as timescales and geoscales do, lets
`library(yourpackage)` alone provide `recast()`:

```{r, eval = FALSE}
#' @importFrom multiscales recast
#' @export
multiscales::recast
```

## Things to keep in mind

* **Never test a class name by spelling.** An S7 class defined in a package is
  `"yourpackage::Industry"` in `class()`, but plain `"Industry"` under
  `pkgload::load_all()`. Use `scale_is()`, or `S7::S7_inherits()` against the
  class object.
* **Stored objects carry their class.** An object saved to `data/` or an `.rds`
  records its class and its parent's. If either changes -- a rename, a new
  parent -- the saved object no longer passes `scale_is()` and every verb
  rejects it with "`x` must be a Scale object". Rebuild shipped datasets after
  any class change.
* **Keep the dimension in the dimension package.** multiscales depends on no
  dimension package; nothing in it should need to know that calendars or
  regions exist.
